---
title: The R package NNT
subtitle: Computation of Number Needed to Treat and confidence interval
author: Gilles Guillot and Cécile Pizot
date: "`r Sys.Date()`"
# output: rmarkdown::html_vignette
abstract: This vignette introduces briefly  the R package NNT for interval estimation of the Number Needed to Treat. A shiny app providing a graphical interfact can be found on [https://i-pri.shinyapps.io/CI_NNT/](https://i-pri.shinyapps.io/CI_NNT/). This is a companion program to the article _Efforts needed for preventing breast and colorectal cancer through changes in dietary patterns_ [@NNT_Mullie_2020].
output:
  #rmarkdown::html_vignette:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
header-includes:
   - \usepackage{setspace}
   - \onehalfspacing
   - \usepackage{manfnt}
bibliography: 
 - /home/gilles/Dropbox/com/bibtex/biblio.bib
 - /home/gilles/Dropbox/com/bibtex/gilles.bib 
# - ./biblio.bib
# - ./gilles.bib
vignette: >
  %\VignetteIndexEntry{Dose-response analysis of mixture data with the model deviation ratio method.}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  background=col2rgb("cornsilk3"),
  highlight = TRUE
)
```

```{r setup, echo=FALSE}
library(MDR)
library(ggplot2)
library(gridExtra)
# library(bookdown)
```

# Method


## Number needed to treat

For a group of $n$ persons following a treatment with associated absolute risk $p_A$, the expected number of cases is $n\times p_A$. 
For the same group of $n$ persons, a change from treatment $A$ with absolute risk $p_A$ to treatment $B$ with associated risk $p_B$ will induce a change in the number of cases of $n \times (p_A - p_B)$. The number of persons needed to treat (NNT) is defined as the inverse of absolute risk reduction and represent the number of persons to treat that will induce a reduction in expected number of cases of one unit. 


## Confidence intervals for the absolute risk reduction

The absolute risk reduction ($ARR$) is defined as the difference $p_A - p_B$. Computing an interval prediction for a difference in proportion is a classical problem in biostatistics. 
 @newcombe1998interval reviewed eleven methods to achieve this task and concluded that a method based on the Wilson score [@wilson1927probable] appears to be the best trade-off between computational convenienence and accuracy. 
We therefore employed the Wilson score method to derive a $(1-\alpha)$  confidence interval   for absolute risk reduction, which brings [reprinted from @bender2001calculating, Eqs 3 and 4]  $CI^{ARR}_{1-\alpha} = [L^{ARR};U^{ARR}]$  with:
$$L^{ARR} = p_A - p_B - \delta \;\;  \mbox{ and }  \;\; U^{ARR} = p_A - p_B + \varepsilon $$
where
$$\delta=\sqrt{(p_A-l_A)^2+(u_B-p_B)^2}  \;\;  \mbox{ and }  \;\; \varepsilon = \sqrt{(u_A-p_A)^2+(p_B-l_B)^2}$$

$$l_i = \varphi_i - \sqrt{\varphi_i^2-\psi_i}   \;\;  \mbox{ and }  \;\;  u_i =\varphi_i + \sqrt{\varphi_i^2-\psi_i} $$
$$\varphi_i =  \frac{2e_i+z_{1-\alpha/2}^2}{2(n_i+z_{1-\alpha/2}^2)}    \;\;  \mbox{ and }  \;\; \psi_i =  \frac{e_i^2}{n_i^2+n_i z_{1-\alpha/2}^2}$$

## Confidence interval for the number needed to treat 

As noted e.g. by @altman1998confidence and @bender2001calculating, if $CI^{ARR}_{1-\alpha} = [L;U]$ is a $(1-\alpha)$ confidence interval for $ARR$, then a $(1-\alpha)$ confidence interval for $NNT$ can be obtained as

\begin{equation}\label{CI1}
CI^{NNT}_{1-\alpha} = [1/U;1/L] \;\; \mbox{ if } \;\; [L;U]  \mbox{ does not contain } 0
\end{equation} and 

\begin{equation}\label{CI2}
CI^{NNT}_{1-\alpha} = (-\infty;1/L] \cup [1/U;+\infty) \mbox{ if } \;\; [L;U]  \mbox{ contains } 0 \, ,
\end{equation}
negative values for NNT being often referred to as number needed to harm [@hutton2000number].

# Package overview

The NNT package consists of a single function called `compute_NNT` taking as input

* `e`: a bivariate numeric vector of counts of events in the two treatment groups
* `n`: a bivariate numeric vector of size of the two treatment groups
* `alpha`: a numeric value between 0 and 1 so as to obtain $1-\alpha$ confidence interval

 It returns a list whose elements include point estimates of ARR, NNT and 1-alpha confidence intervals.

# Coverage study


To further assess the accuracy of the Wilson score method, we carried out a small simulation study as follows: 
We drew Monte Carlo simulations of pairs of cases $(e_A,e_B)$ from binomial distributions $B(n_A,p_A)$ and $B(n_B,p_B)$. 
For each simulated pair $(e_A,e_B)$, we computed the confidence interval for the NNT using the Wilson score method. 
Replicating this procedure 1000 times, we computed the coverage, namely the proportion of confidence intervals containing the true risk reduction, cf code below. 
We compared the  coverage to the nominal level of the confidence intervals on various sets of risk and various sets of nominal levels.


```{r,echo=FALSE}
source('../R/compute_NNT.R')
```

```{r}
cover = function(p1,p2,n,nrep,alpha)
{
  nnt = 1/(p1-p2)
  ncov=0
  for(irep in 1:nrep)
  {
    e = rbinom(n=2,p=c(p1,p2),size=n)
    res_ci = compute_NNT(e,n,alpha = alpha)
    if(!is.na(res_ci$LL_NNT))
    { 
      if( (nnt > res_ci$LL_NNT) & (nnt < res_ci$UL_NNT))  ncov = ncov+1
    }else
    {
      if( (nnt < res_ci$UL_NNT_neg) | (nnt > res_ci$LL_NNT_pos)) ncov = ncov+1
    }
  }
  ncov/nrep
}
```



We get the table below, showing that  for values investigated, the method seems quite accurate. Note that in the last two rows, $p_A$ and $p_B$ are  picked 
from actual data reported by @farvid2016fruit.

$p_A$ | $p_B$ | NNT              | $n_A$ $n_B$   | $1-\alpha$ | Coverage
-----|------|------------------|-------------|----------|----------
1/5 | 1/10  | `r 1/(1/5-1/10)` | 1000        | 0.95     | `r signif(cover(p1=1/5,p2=1/10,n=rep(1000,2),nrep=10000,alpha=0.05),dig=3)`
1/10| 1/5   | `r 1/(1/10-1/5)` | 1000        | 0.95     | `r signif(cover(p1=1/10,p2=1/5,n=rep(1000,2),nrep=10000,alpha=0.05),dig=3)`
1/5 | 1/10  | `r 1/(1/5-1/10)` | 1000        | 0.99     | `r signif(cover(p1=1/5,p2=1/10,n=rep(1000,2),nrep=10000,alpha=0.01),dig=3)`
1/10| 1/5   | `r 1/(1/10-1/5)` | 1000        | 0.99     | `r signif(cover(p1=1/10,p2=1/5,n=rep(1000,2),nrep=10000,alpha=0.01),dig=3)`
1/5 | 1/10  | `r 1/(1/5-1/10)` | 1000        | 0.9      | `r signif(cover(p1=1/5,p2=1/10,n=rep(1000,2),nrep=10000,alpha=0.1),dig=3)`
1/5 | 1/10  | `r 1/(1/5-1/10)` | 100         | 0.95     | `r signif(cover(p1=1/5,p2=1/10,n=rep(1000,2),nrep=10000,alpha=0.05),dig=3)`
1/5 | 1/10  | `r 1/(1/5-1/10)` | 50          | 0.95     | `r signif(cover(p1=1/5,p2=1/10,n=rep(1000,2),nrep=10000,alpha=0.05),dig=3)`
1/1000 | 1/2000  | `r 1/(1/1000-1/2000)` | 10000        | 0.95     | `r signif(cover(p1=1/10,p2=2/10,n=rep(1000,2),nrep=10000,alpha=0.05),dig=3)`
`r signif(315/134534,dig=3)` | `r signif(263/134375,dig=3)` | `r round(1/(315/134534 - 263/134375))` | 134535 134375 | 0.95 | `r signif(cover(p1=315/134534,p2=263/134375,n=c(134535,134375),nrep=10000,alpha=0.05),dig=3)`
`r signif(236/134375,dig=3)` | `r signif(250/134590,dig=3)` | 10028 | 134375 134590 | 0.95 | `r signif(cover(p1=315/134375,p2=263/134590,n=c(134375,134590),nrep=10000,alpha=0.05),dig=3)`




# References
